plyranges: a grammar for transforming genomics data

### 2018-04-17 paper.Rmd
fix some typos

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use an 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
Query Language (GMQL) implements a relational algebra for combining genomic datasets 
@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

@@

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behaviour that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
defines an API that maps notions from the general relational algebra to 
operations on tidy data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.

@@

The `plyranges` DSL is built on the most general Bioconductor data structure, 
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example a GRanges can be used to represent a gene with
its chromosomal coordinates, and along with its exon identifiers; or 
a GRanges can represent an RNA-seq experiment as a set of genes
with a corresponding count column.

By definition the GRanges follow  the tidy data pattern: it is a rectangular 
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation. Hence, we have designed the `plyranges` DSL to extend the grammar
and design principles of `dplyr`:  cohesion, consistency, endomorphism,
and fluency. All of these principles are defined and discussed
below in the context of the GRanges class. Where applicable we contrast our 
design to the existing Bioconductor infrastructure.

>>>>

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use an 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric 
Query Language (GMQL) implements a relational algebra for combining genomic datasets 
@Kaitoua2017-pw.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

@@

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending on the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behaviour that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined as a tabular
data structure that has observations as rows and columns as variables,
and all measurements pertain to a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relate to the design of an experiment
and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
defines an API that maps notions from the general relational algebra to 
operations on tidy data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.

@@

The `plyranges` DSL is built on the most general Bioconductor data structure, 
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example, a GRanges can represent a gene with
its chromosomal coordinates, exonic structure, and a count column
summarized from an RNA-seq experiment.

By definition GRanges follows the tidy data pattern: it is a rectangular 
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation. Hence, we have designed the `plyranges` DSL to extend the grammar
and design principles of `dplyr`:  cohesion, consistency, endomorphism,
and fluency. All of these principles are defined and discussed
below in the context of the GRanges class. Where applicable we contrast our 
design to the existing Bioconductor infrastructure.
